Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2017-10-01 output/breed_paper.Rnw
Adds secondary analysis

@@

California (CA) rice production represents one such production
system. Rice yields in CA during the period from 1920 -- 1990
increased rapidly (Fig. \ref{fig:state_yield}), in part due to
improvements in rice genetics (i.e., the ``Green Revolution'') through
the efforts of the California Cooperative Rice Research Foundation
(CCRRF; a collaboration between the University of California, the
USDA-Agricultural Research Service, and farmer-funded
research). However, since 1990 the rate of increase has slowed
(Fig. \ref{fig:state_yield}) despite continuous crop improvement
\citep{ccrrf}. For the past 30 years, each newly developed cultivar
has demonstrated a yield advantage over the previous cultivars in the
Statewide Variety Trials (an annual multi-site variety trial program
that compares experimental entries side-by-side with ``check''
cultivars). However, in a seeming contradiction these same reports
show stagnant yields across release announcements (i.e., reports
accompanying the official public release) from the previous years. For
example, when cultivar M-202 was released in 1985, it was reported to
yield an average of 11 Mg ha\tsups{-1} \citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1} \citep]14\%
MC;][]{m209}. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed \citep{Peng1998, Peng2010,
Peng2000, Datta1995}, often termed yield erosion \citep{Peng1998}. It
has been speculated that this yield erosion is due to the inability of
a cultivar to adapt to changing biotic and abiotic conditions. To
better understand how plant improvement and yield erosion are
contributing to the improvement of rice yield in farmers' fields, we
teseted the hypotheses: either (1) the genetic yield advantage
increased over the last 30+ years, or (2) the genetic yield advantage
has not increased over the last 30+ because the erosion of yield over
time creates an illusion of improved yields.

@@

The CCRRF Rice Experiment Station (RES) is located north-west of
Biggs, CA (39.4648, -121.7342), and has been the central location for
CCRRF's efforts to develop improved rice varieties adapted to CA since
1969. The climate at the RES is Mediterranean, characterized by mild
winters during which most of the annual precipitation occurs
(\Sexpr{18*24.5} to \Sexpr{25*24.5} mm annually) and warm summers
largely free of precipitation events. Soils at the RES are classified
as Esquon-Neerdobe clays with roughly 2\% soil organic matter
\citep{ssurgo} in the top 15cm. As the central location for the CCRRF
breeding program, the RES is researcher-managed (as opposed to
farmer-managed trials) and plants the majority of experimental plots for the
Statewide Variety Trials (approximately 3 to 4 times the number of
experimental plots compared to any other location). To avoid potential
complications due to site, climate, and management differences at the
other 6 to 8 farmer-managed trials per year we focused our
investigation on the RES.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as these cultivars constitute approximately
90\% of the planted rice area in CA. These cultivars are typcically
evaluated in the Statewide Variety Trials for 3 to 5 years prior to
their official release. For these cultivars planted in breeding trials
at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations (replicated
3--4 times in a CRB design) were collected leading to a final data set
of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
representing \Sexpr{length(unique(mod_data[,'id']))} cultivars,
including every publicly released medium grain variety developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of public release and the time elapsed (in years) since release was
calculated.

@@

To test each hypothesis, the influence of breeding and yield erosion
were quantified on yield over time. Yield was modeled as a response to
the number of years since the variety was officially released and the
year of release using a mixed-effects linear regression. To account
for similarities between years and between cultivars, year and
cultivar were included in the model as random effects.

@@

Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
average yield for the RES across varieties and years, $Rate_{yield\
erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
(negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
year, and $\beta_{cultivar}$ is the random effect of cultivar. For
comparison to estimates where yield erosion is not taken into account,
the model was used as described above except omitting the $Rate_{yield\ erosion\ i}$ term.

@@

\caption{The experimental yield of medium grain rice varieties for the
  two years prior to each cultivar's public release. Points
($\bullet$) and bars (|) represent the mean and standard error of the
raw data, respectively. The dashed line and shaded area are the
genetic yield improvement and credible interval as estimated by the model.}
\label{fig:first_yrs}
\end{figure}

@@

@
\caption{The relationship between yield and the years since release
  for medium grain rice varieties. The estimated erosion of yield over time
  (averaged over differences in cultivar and year) is estimated to be
  \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}. Points ($\bullet$)
  and bars (|) represent the mean and standard error of the raw data,
  respectively. The shaded area is the 90\% credible interval.}
% For each year, point size by n (open circles)
\label{fig:RES_year}
\end{figure}

@@

To be clear, our investigation here is narrowly focused on the
rough-rice/field grain yield. Other advancements due to plant
improvement, including increased quality \citep{mckenzie1993} and
milling yield \citep{mckenzie1994}, have been made over the study
period and are important. Also, crop duration has been reduced. For
example, during this study period, M-202 and M-206 have been the
predominate varieties grown in CA, and while their yield potentials
were similar at the time of release (Fig. 5), M-206 matures about 6
days earlier than M-202 (Table 2). Shorter duration is very important
in CA due to an already short growing season and water limitations
(shorter duration varieties would presumably require less irrigation
water).

>>>>

\address[ucce]{University of California Cooperative Extension,
Colusa, CA 95932, USA}

@@

California (CA) rice production represents one such production
system. Production in CA is predominately focused on premium quality
medium grain \textit{japonica} cultivars (e.g., CalRose rice), and
rice from CA is recognized globally for its quality. Rice yields in CA
increased rapidly during the period from 1920 -- 1990
(Fig. \ref{fig:state_yield}), in part due to improvements in rice
genetics (i.e., the ``Green Revolution''). Most cultivars in use in CA
are developed by the California Cooperative Rice Research Foundation
(CCRRF; a collaboration between the University of California, the
USDA-Agricultural Research Service, and farmer-funded
research). Although grain quality has been maintained or even improved, rate of yield
increase has slowed since the 1990s (Fig. \ref{fig:state_yield})
despite continuous crop improvement \citep{ccrrf}. For the past 30
years, each newly developed cultivar has demonstrated a yield
advantage over the previous cultivars in the Statewide Variety Trials
(an annual multi-site variety trial program that compares experimental
entries side-by-side with ``check'' cultivars). However, in a seeming
contradiction these same reports show stagnant yields across release
announcements (i.e., reports accompanying the official public release)
from the previous years. For example, when cultivar M-202 was released
in 1985, it was reported to yield an average of 11 Mg ha\tsups{-1}
\citep[14\% moisture content (MC);][]{m202}. In 2015, the reported
average yield of the newest released cultivar M-209 was 10.8 Mg
ha\tsups{-1} \citep]14\% MC;][]{m209}. In other rice systems, a
decrease in a cultivar's yield performance over time has been observed
\citep{Peng1998, Peng2010, Peng2000, Datta1995}, often termed yield
erosion \citep{Peng1998}. It has been speculated that this yield
erosion is due to the inability of a cultivar to adapt to changing
biotic and abiotic conditions. To better understand how plant
improvement and yield erosion are contributing to the improvement of
rice yield in farmers' fields, we teseted the hypotheses: either (1)
the genetic yield advantage increased over the last 30+ years, or (2)
the genetic yield advantage has not increased over the last 30+
because the erosion of yield over time creates an illusion of improved
yields.

@@

The CCRRF Rice Experiment Station (RES) is located north-west of
Biggs, CA (39.4648, -121.7342), and has been the central location for
CCRRF's efforts to develop improved rice cultivars adapted to CA since
1969. The climate at the RES is Mediterranean, characterized by mild
winters during which most of the annual precipitation occurs
(\Sexpr{18*24.5} to \Sexpr{25*24.5} mm annually) and warm summers
largely free of precipitation events. Soils at the RES are classified
as Esquon-Neerdobe clays with roughly 2\% soil organic matter
\citep{ssurgo} in the top 15cm. As the central location for the CCRRF
breeding program, the RES is researcher-managed (as opposed to
farmer-managed trials) and plants the majority of experimental plots for the
Statewide Variety Trials (approximately 3 to 4 times the number of
experimental plots compared to any other location). To avoid potential
complications due to site, climate, and management differences at the
other 6 to 8 farmer-managed trials per year we focused our
investigation on the RES.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as these cultivars constitute approximately
90\% of the planted rice area in CA. These cultivars are typcically
evaluated in the Statewide Variety Trials for 3 to 5 years prior to
their official release. For these cultivars planted in breeding trials
at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations (replicated
3--4 times in a CRB design) were collected leading to a final data set
of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
representing \Sexpr{length(unique(mod_data[,'id']))} cultivars,
including every publicly released medium grain cultivar developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of public release and the time elapsed (in years) since release was
calculated.

@@

To test each hypothesis, the influence of breeding and yield erosion
were quantified on yield over time. Yield was modeled as a response to
the number of years since the cultivar was officially released and the
year of release using a mixed-effects linear regression. To account
for similarities between years and between cultivars, year and
cultivar were included in the model as random effects.

@@

Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
average yield for the RES across cultivars and years, $Rate_{yield\
erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
(negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
year, and $\beta_{cultivar}$ is the random effect of cultivar. For
comparison to estimates where yield erosion is not taken into account,
the model was used as described above except omitting the $Rate_{yield\ erosion\ i}$ term.

@@

\caption{The experimental yield of medium grain rice cultivars for the
  two years prior to each cultivar's public release. Points
($\bullet$) and bars (|) represent the mean and standard error of the
raw data, respectively. The dashed line and shaded area are the
genetic yield improvement and credible interval as estimated by the model.}
\label{fig:first_yrs}
\end{figure}

@@

@
\caption{The relationship between yield and the years since release
  for medium grain rice cultivars. The estimated erosion of yield over time
  (averaged over differences in cultivar and year) is estimated to be
  \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}. Points ($\bullet$)
  and bars (|) represent the mean and standard error of the raw data,
  respectively. The shaded area is the 90\% credible interval.}
% For each year, point size by n (open circles)
\label{fig:RES_year}
\end{figure}

@@

To be clear, our investigation here is narrowly focused on the
rough-rice/field grain yield. Other advancements due to plant
improvement, including increased quality \citep{mckenzie1993} and
milling yield \citep{mckenzie1994}, have been made over the study
period and are important. Also, crop duration has been reduced. For
example, during this study period, M-202 and M-206 have been the
predominate cultivars grown in CA, and while their yield potentials
were similar at the time of release (Fig. 5), M-206 matures about 6
days earlier than M-202 (Table 2). Shorter duration is very important
in CA due to an already short growing season and water limitations
(shorter duration cultivars would presumably require less irrigation
water).
